\large{\bf Looking forwards and backwards through locally regulated populations }

### 2022-05-09 forwards-and-backwards.tex
edits 3

Theorem~\ref{thm:nonlocal_convergence} shows convergence to the solution to a nonlocal PDE
of the form \eqref{eqn:nonlocal_pde}.
Since the coefficients at $x$ in \eqref{eqn:nonlocal_pde}
depend only on the region nearby to $x$,
we would often expect solutions of the nonlocal PDE to be close to the corresponding local PDE
as the interaction distances $\epsilon_r$, $\epsilon_\gamma$ and $\epsilon_F$
go to zero.
The following propositions give two concrete situations in which that is true.

\comment{I've not got straight exactly how we're framing this
    (namely, does $\eta$ have a density?).
    Also, is this one or two propositions --
    we said initially this would be two,
    one for FKPP and one for PME,
    but then maybe the argument was more general?
}

\begin{proposition}
    \label{prop:nonlocal_to_local}
    Suppose that $\eta_\epsilon(t,dx)$ is a solution to equation \eqref{eqn:nonlocal_pde}
    with $r(x, m) = 1$ for all $x$ and $m$
    and some $\epsilon_r < \epsilon_\gamma = \epsilon_F = \epsilon$
    \comment{and assumptions on $\gamma$? and $F$?}.
    Then $\eta_\epsilon(t, dx)$ converges \comment{in what sense?} as $\epsilon \to 0$
    to $u(t, x) dx$, which solves
    \begin{align} \label{eqn:PDE}
        \partial_t u(t, x)
        =
        \DG^* \left( \gamma(\cdot, u(t, \cdot) u(t, \cdot)  \right)(x)
        + F(x, u(t, x)) u(t, x) .
    \end{align}
\end{proposition}

Theorem \ref{thm:nonlocal_convergence} combined with Proposition \ref{prop:nonlocal_to_local}
imply that we can take the $N \to \infty$ limit
followed by $\epsilon \to 0$
to obtain solutions to the PDE \eqref{eqn:PDE}.
However, it is of substantial interest to know whether
we can take those two limit simultaneously.
The general case seems difficult,
but we can prove such ``diagonal'' convergence in the following situation.

\begin{theorem}[Convergence to a PDE]
    \label{thm:local_convergence}
    Let $(\eta^\theta_t)_{t \geq 0}$
    be as defined in Definition \ref{def:model_setup}
    and assume that as $N \to \infty$, $N \to \infty$
    and $\epsilon_r < \epsilon_\gamma = \epsilon_F = \epsilon \to 0$
    in such a way that $\theta/N \to 0$
    and
    \comment{TODO: some conditions on $\epsilon$ and $r$ and $\gamma$ and $F$}.
    Then the sequence of stochastic processes $(\eta^\theta_t)_{t \ge 0}$
    converges \comment{in some sense}
    to a measure-valued process with a density $\varphi(t, x)$
    that solves
    \begin{align}
        \partial_t \varphi(t, x)
        &=
        \text{whatever the final form is here} .
    \end{align}
\end{theorem}

@@

Now that we have established what we can say about how population density changes with time,
we turn to results on ancestral lineages,
i.e., how genealogical ancestry moves across the landscape.
Informally,
a \emph{lineage} $(L_t)_{t \ge 0}$
begun at spatial location $L_0 = x$
can be obtained by picking a focal individual uniformly at $x$,
and then for each amount of time $t$ in the past,
setting $L_t$ to the spatial location of the individual from whom
the focal individual inherits.
Since in our model individuals have only one parent, this is unambiguous.
Although we did not explicitly retain such information,
it is clear that
for finite $\theta$ (and hence, finite $N$),
one could construct the distribution of a lineage $(L_t)_{t=0}^T$
given the history of the population $(\eta^\theta_t)_{t = 0}^T$,
for each starting location to which $\eta^\theta_T$ assigns positive mass.
It is less clear, however, how to formally retain such information in the limit.
The \emph{lookdown construction} in Section~\ref{sec:lookdown}
provides just such a construction,
where lineages are retained in the $N \to \infty$ limit.
Roughly speaking,
this works by assigning to each particle a unique ``level''
that functions as a label and thus allows reconstruction of lineages,
but also (in some sense) orders individuals by eventual reproductive output,
allowing a formal limit.
See~\ref{etheridge/kurtz:2018} for an introduction to these ideas.

A lineage is, in general, a well-defined stochastic process,
but there are two important questions
that affect its tractability.
First, when is it a Markov process given the population process?
(I.e., given $(\eta_t)_{t=0^T}$ that records numbers of individuals
but not their ancestry.)
We expect the answer here to be that a lineage is Markov
only in the $N \to \infty$ limit.
\comment{(or, do we know this? we can cite Barton, Depaulis and Etheridge at least for discussion.)}
Second, does the process have a tractable description?
If the population process has a density,
then it turns out that a lineage is a diffusion driven by the density.
The remaining case is the superprocess limit ($\theta/N \to \alpha > 0$);
in this case, we expect that a lineage could still be described as a generalized diffusion,
but moving within a stochastic landscape that is singular in $d \ge 2$.
We focus on the more tractable situation where the population process is deterministic.
However, the results here apply to either the local or nonlocal case
(i.e., where the density solves a local or nonlocal PDE).

\comment{
    TODO: make notation agree with that in the proof.
}

\begin{definition}[Generator of a lineage] \label{def:lineage_generator}
    Let $(\varphi(x, t))_{0 \le t \le T}$
    denote the density of a population that solves \eqref{eqn:nonlocal_pde},
    and let $y$ be a point with $\varphi(y, T) > 0$.
    We define $(L_s(y))_{s=0}^T$,
    the ancestral lineage of an individual sampled at $y$ at time $T$,
    to be the position of the unique ancestor of $y$ alive at time $T - s$.
    We define
    $(Q_s)_{s \geq 0}$
    to be the time inhomogeneous semi-group satisfying
    \begin{align*}
        Q_s f(y) := \IE_y[ f(L_s) ] ,
    \end{align*}
    and $\Lgen_s$ to be the generator of this semi-group.
\end{definition}

Our main result identifies a lineage as a diffusion
by characterizing its generator.
To avoid complexity,
we do this in the case where dispersal is $N(\meanq/\theta, \sigma^2 I/\theta)$,
in which case $\DG = \sigma^2 \Delta + \meanq \cdot \grad$.

\comment{Make forwards-backwards time notation reasonable in this.}

\comment{Perhaps write generator as depending on $\phi$ rather than $s$?
    Then we can say that the generator is $\Lgen_{\varphi(T-s)}$.}

\begin{theorem} \label{thm:lineages}
    In the context of Definition~\ref{def:lineage_generator},
    \begin{align}
        \label{eqn:lineage_generator}
        \Lgen_s f
        &=
        \frac{r}{\varphi_{T-s}}
        \left[
            \DG^*(\gamma \varphi_{T-s} f) 
            - f \DG^* (\gamma \varphi_{T-s})
        \right] \\
        &= \label{eqn:lineage_generator2}
        r\gamma
        \left[
            \sum_{ij} \covq_{ij} \partial_{ij} f
            + \sum_j \vec{m}(s)_j \partial_j f
        \right] ,
    \end{align}
    where $\vec{m}$ is the vector
    $$
    \vec{m}(s)_j
    =
    2 \sum_i C_{ij} \partial_i \log(\gamma \varphi_{T-s})
    + 2 \sum_i \partial_i C_{ij}
    - \meanq_j .
    $$
\end{theorem}

To make the generator readable, we've written it in concise notation,
omitting the dependencies on location and population density,
which itself changes with time.
When interpreting this,
remember that everything depends on location and density at that location and time --
for instance, ``$r$'' is actually $r(x, \varphi(T-s, x))$.

\begin{corollary} \label{cor:lineages_simple}
    In the context of Definition~\ref{def:lineage_generator},
    if the dispersal process is isotropic (i.e., $\covq = \sigma^2 I$),
    then
    \begin{equation}
        \Lgen_s f
        =
        \sigma^2 r \gamma
        \left(
            \Delta f
            +
            \left(
                2 \grad \log(\gamma \varphi_s)
                - \meanq
            \right)
            \cdot \grad f
        \right) .
    \end{equation}
    In other words, 
    a lineage behaves as a diffusion
    that is driven by a Brownian motion 
    run at speed $\sigma^2$ multiplied by the local per-capita production of offspring ($r \gamma$)
    in a potential equal to the local birth output tilted by migraiton bias
    ($\varphi \gamma \exp(-\meanq \cdot x / 2 \sigma^2)$).
\end{corollary}

\begin{corollary} \label{cor:wavefront}
    In addition to the assumptions of Corollary~\ref{cor:lineages_simple},
    suppose that the population process is described by a traveling wave with speed $\wavespeed$,
    i.e., the population has density
    $\varphi(t, x) = w(x - t \wavespeed)$
    where $w$ solves
    \begin{align*}
        r \Delta (\gamma w) + w F(w) + \wavespeed \cdot \grad w = 0 .
    \end{align*}
    Then the generator $\Lgen$
    of the motion of a linages in this population is
    \begin{align}
        \Lgen f
        &=
        r \gamma
        \left(
            \Delta f
            +
            2 \grad \log (\gamma w)
            \cdot \grad f
        \right)
        + \wavespeed \cdot \grad f .
    \end{align}
\end{corollary}

>>>>

Although the coefficients at $x$ in \eqref{eqn:nonlocal_pde} are nonlocal,
they depend only on the region nearby to $x$,
and so we would expect solutions of the nonlocal PDE to be close to the corresponding local PDE
as the interaction distances $\epsilon_r$, $\epsilon_\gamma$ and $\epsilon_F$
go to zero:
for instance, if $\eta_t$ has density $\varphi_t$
then $\gamma(x, \smooth{\gamma} \eta_t(x)) \to \gamma(x, \varphi_t(x))$ as $\epsilon_\gamma \to 0$.
The following propositions give two concrete situations in which that is true.

\begin{proposition}
    \label{prop:nonlocal_to_local}
    Suppose that $\eta^\epsilon_t$ is a weak solution to equation \eqref{eqn:nonlocal_pde}
    with $r(x, m) = 1$ for all $x$ and $m$
    and some $\epsilon_r + 1/\theta < \epsilon_\gamma = \epsilon_F = \epsilon$
    \comment{and assumptions on $\gamma$? and $F$?}.
    Then $\eta^\epsilon_t(dx)$ converges \comment{in what sense?} as $\epsilon \to 0$
    to $\varphi_t(x) dx$, which is a (strong) solution to
    \begin{align} \label{eqn:PDE}
        \partial_t \varphi(t, x)
        =
        \DG^* \left( \gamma(\cdot, \varphi(t, \cdot)) \varphi(t, \cdot)  \right)(x)
        + F(x, \varphi(t, x)) \varphi(t, x) .
    \end{align}
\end{proposition}

Theorem \ref{thm:nonlocal_convergence} combined with Proposition \ref{prop:nonlocal_to_local}
imply that we can take the $N \to \infty$ limit
followed by $\epsilon \to 0$
to obtain solutions to the PDE \eqref{eqn:PDE}.
However, it is of substantial interest to know whether
we can take those two limits simultaneously.
The general case seems difficult,
but we can prove such ``diagonal'' convergence in the following situation.

\begin{theorem}[Convergence to a PDE]
    \label{thm:local_convergence}
    Let $(\eta^\theta_t)_{t \geq 0}$
    be as defined in Definition \ref{def:model_setup}
    and assume that as $N \to \infty$, $\theta \to \infty$
    and $\epsilon_r + 1/\theta < \epsilon_\gamma = \epsilon_F = \epsilon \to 0$
    in such a way that $\theta/N \to 0$
    and
    \comment{TODO: some conditions on $\epsilon$ and $r$ and $\gamma$ and $F$}.
    Then the sequence of stochastic processes $(\eta^\theta_t)_{t \ge 0}$
    converges \comment{in some sense}
    to a measure-valued process with a density $\varphi(t, x)$
    that solves
    \begin{align}
        \partial_t \varphi(t, x)
        &=
        \text{whatever the final form is here} .
    \end{align}
\end{theorem}

@@

Now that we have established what we can say about how population density changes with time,
we turn to results on ancestral lineages,
i.e., how genealogical ancestry can be traced back across the landscape.
Informally,
a \emph{lineage} $(L_t)_{t \ge 0}$
begun at spatial location $L_0 = x$
can be obtained by picking a focal individual uniformly at $x$,
and then for each amount of time $t$ in the past,
setting $L_t$ to the spatial location of the individual from whom
the focal individual inherits at that time in the past.
Since in our model individuals have only one parent, this is unambiguous.
Although we did not explicitly retain such information,
it is clear that for finite $N$
one could construct the distribution of a lineage $(L_t)_{t=0}^T$
given the history of the population $(\eta^N_t)_{t = 0}^T$,
for each starting location to which $\eta^N_T$ assigns positive mass.
It is less clear, however, how to formally retain such information in the limit.
The \emph{lookdown construction} in Section~\ref{sec:lookdown}
provides just such a construction,
where lineages are retained in the $N \to \infty$ limit.
Roughly speaking,
this works by assigning to each particle a unique ``level''
that functions as a label and thus allows reconstruction of lineages,
but also (in some sense) orders individuals by eventual reproductive output,
allowing a formal limit.
See~\ref{etheridge/kurtz:2018} for an introduction to these ideas.

A lineage is, in general, a well-defined stochastic process,
but there are two important questions
that affect its tractability.
First, when is it a Markov process given the population process?
(In other words, given $(\eta_t)_{t=0}^T$ that records numbers of individuals
but not their ancestry.)
We expect the answer here to be that a lineage is Markov
only in the $N \to \infty$ limit
(see \citet{barton/depaulis/etheridge:2002} for discussion).
Second, does the process have a tractable description?
If the population process has a density,
then it turns out that a lineage is a diffusion driven by the density.
The remaining case is the superprocess limit ($\theta/N \to \alpha > 0$);
in this case, we expect that a lineage could still be described as a generalized diffusion,
but moving within a stochastic landscape that is singular in $d \ge 2$.
We focus on the more tractable situation where the population process is deterministic.
However, the results here apply to either the local or nonlocal case
(i.e., where the density solves a local or nonlocal PDE).

\begin{definition}[Generator of a lineage] \label{def:lineage_generator}
    Let $(\varphi_t(x))_{0 \le t \le T}$
    denote the density of a population that solves \eqref{eqn:nonlocal_pde},
    and let $y$ be a point with $\varphi_T(y) > 0$.
    We define $(L_s(y))_{s=0}^T$,
    the ancestral lineage in this population of an individual sampled at $y$ at time $T$,
    to be the position of the unique ancestor of $y$ alive at time $T - s$.
    We define
    $(Q_s)_{s \geq 0}$
    to be the time inhomogeneous semi-group satisfying
    \begin{align*}
        Q_s f(y) := \IE_y[ f(L_s) ] .
    \end{align*}
\end{definition}

Our main result identifies a lineage as a diffusion
by characterizing its generator.

\begin{theorem} \label{thm:lineages}
    For $\varphi: \IR^d \to \IR$, define
    \begin{align}
        \label{eqn:lineage_generator}
        \Lgen_\varphi f
        &=
        \frac{r}{\varphi}
        \left[
            \DG^*(\gamma \varphi f) 
            - f \DG^* (\gamma \varphi)
        \right] \\
        &= \label{eqn:lineage_generator2}
        r\gamma
        \left[
            \sum_{ij} \covq_{ij} \partial_{ij} f
            + \sum_j \vec{m}(s)_j \partial_j f
        \right] ,
    \end{align}
    where $\vec{m}$ is the vector
    $$
    \vec{m}(s)_j
    =
    2 \sum_i C_{ij} \partial_i \log(\gamma \varphi)
    + 2 \sum_i \partial_i C_{ij}
    - \meanq_j .
    $$
    Then $\Lgen_{\varphi_{T-s}}$ is the generator of the semigroup $Q_s$
    of Definition~\ref{def:lineage_generator}.
\end{theorem}

As usual,
to make the generator readable, we've written it in concise notation,
omitting the dependencies on location and population density,
which itself changes with time.
When interpreting this,
remember that everything depends on location and density at that location and time --
for instance, ``$r$'' is actually $r(x, \varphi(x))$.

\begin{corollary} \label{cor:lineages_simple}
    In addition to the assumptions of Theorem~\ref{thm:lineages},
    if the dispersal process is isotropic (i.e., $\covq = \sigma^2 I$),
    then
    \begin{equation}
        \Lgen_\varphi f
        =
        \sigma^2 r \gamma
        \left(
            \Delta f
            +
            \left(
                2 \grad \log(\gamma \varphi)
                - \meanq
            \right)
            \cdot \grad f
        \right) .
    \end{equation}
    In other words, 
    a lineage behaves as a diffusion
    that is driven by a Brownian motion 
    run at speed $\sigma^2$ multiplied by the local per-capita production of offspring ($r \gamma$)
    in a potential equal to the local birth output tilted by migration bias
    ($\varphi_s \gamma \exp(-\meanq \cdot x / 2 \sigma^2)$).
\end{corollary}

\begin{corollary} \label{cor:wavefront}
    In addition to the assumptions of Corollary~\ref{cor:lineages_simple},
    suppose that the population process is described by a traveling wave with speed $\wavespeed$,
    i.e., the population has density
    $\varphi(t, x) = w(x - t \wavespeed)$
    where $w$ solves
    \begin{align*}
        r \Delta (\gamma w) + w F + \wavespeed \cdot \grad w = 0 .
    \end{align*}
    Then the semigroup 
    of the motion of a linages in this population 
    $Q_s$ is time-homogeneous with generator
    \begin{align}
        \Lgen f
        &=
        \sigma^2 r \gamma
        \left(
            \Delta f
            +
            2 \grad \log (\gamma w)
            \cdot \grad f
        \right)
        + (\wavespeed - \meanq) \cdot \grad f .
    \end{align}
\end{corollary}
